---
title: "Divisive clustering of Histogram Data"
author: "Marie Chavent"
date: "27/07/2021"
output:
  html_document:
  toc: yes
highlight: espresso
theme: journal
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = TRUE, cache = TRUE, warning = F)
```



- The crimes dataset concerns official data on violent crime occurence in the USA, together with some social indicators (see the [Communities and Crime Data Set](https://archive.ics.uci.edu/ml/datasets/communities+and+crime)  on the UCI ML Repository). The data, recorded per county, were aggregated with the **HistDAWass** R package at state level in the form of histogram-valued data, . 

```{r}
load("crimes_hist.rda")
```

Here we analyse a set of $n=30$ states with $p=4$ histogram-valued variables where four intervals per histogram are defined by the quartiles of the observed data.

```{r}
library(HistDAWass)
get.MatH.main.info(CRIMES)
get.cell.MatH(CRIMES,1,1) # histogram value of "AL" on "pctLowEdu
plot(CRIMES, type = "HISTO")
```

- The Mallows distance (L2 Wasserstein distance) between the 30 states is calculated with the function  **WH_MAT_DIST** of the R package **HistDAWass**. The histogram-valued data are standardized according to the procedure available in this function.

```{r}
DMAT <- WH_MAT_DIST(x = CRIMES, standardize = TRUE)
```


- The divisive clustering of the 30 states is performed with the **div_diss** function based on the **divclust** function of the **divclust** R package.


```{r}
source("div_diss.R")
source("choose_question_diss.R")
source("inertdiss-func.R")
```

```{r}
# Matrix of the medians values
M <- as.matrix(get.MatH.stats(CRIMES, stat = "median")$mat)
```

The **div_diss** function takes two values in input :

  - The matrix **M**  of the median values used to build the binary questions.
  - The matrix **DMAT** of dissimilarities used to perform the criterion.
  

```{r, fig.width=15, fig.height=7}
# divisive clustering
library(divclust)
tree <- div_diss(data=M,as.dist(DMAT)) 
plot(tree, nqbin = 4)
```

- The number of clusters is chosen with the Silhouette Index.

```{r}
kmax <- 10
si <- rep(NA, kmax-1)
for (k in 2:kmax)
{
  partdiv <- div_diss(data=M,as.dist(DMAT), K=k)$which_cluster
  sidiv <- cluster::silhouette(partdiv, as.dist(DMAT))
  si[k-1] <- summary(sidiv)$avg.width
}
names(si) <- paste("k=",2:10, sep="")
```

```{r}
barplot(si, col=4,
        xlab="number of clusters", ylab="Silhouette indice")
```

The maximum value is attained for $k = 2$, then we observe a local maximum
for $k = 5$. The partition in two clusters provides an explained inertia (ratio
of between to total sum of squares) of just 32.9%, whereas the partition in
five clusters explains 58.8% of the total inertia. 

```{r}
# explained inertia 
div_diss(data=M,as.dist(DMAT), K=2)$B #for k=2
div_diss(data=M,as.dist(DMAT), K=5)$B  #for k=5
```


- We therefore retain the partition in $k=5$ clusters.

```{r}
tree <- div_diss(data=M,as.dist(DMAT), K=5) 
tree$clusters
tree$description
```

```{r}
partdiv <- tree$which_cluster
```

- Values of the Silhouette Index for the partitios in k = 5 clusters.
```{r}
sidiv <- cluster::silhouette(partdiv, as.dist(DMAT))
rownames(sidiv) <- rownames(DMAT)
plot(sidiv, cex.names = 0.7, main="", 
     sub="", xlab="Silhouette width", border=1,
     do.clus.stat=FALSE, do.n.k=FALSE,
     col = c("red", "green", "blue", "purple", "pink"))
```


- Ward hierarchical clustering with Mallows distance (L2 Wasserstein distance).


```{r}
n <- get.MatH.nrows(CRIMES)
tree <- hclust(as.dist(DMAT^2/(2*n)), method = "ward.D")
plot(tree, hang=-1, sub="", xlab="", main="Ward dendrogram with Mallows distance") 
rect.hclust(tree, k=5)
sum(tree$height[26:29])/ sum(tree$height) # 58.8 % of explained inertia
```

The 5-clusters partition of Ward explains 58.8 % of inertia i.e. only 2% more than divclust but withtout monothetic description.